\documentclass[12pt, a4paper]{article}

% ── Encoding & fonts ──────────────────────────────────────────────────────────
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}

% ── Page layout ───────────────────────────────────────────────────────────────
\usepackage[
    top=2.5cm, bottom=2.5cm,
    left=3cm,  right=2.5cm
]{geometry}
\usepackage{setspace}
\onehalfspacing

% ── Maths ─────────────────────────────────────────────────────────────────────
\usepackage{amsmath}
\usepackage{amssymb}

% ── Figures & tables ──────────────────────────────────────────────────────────
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{float}
\usepackage{multirow}

% ── Hyperlinks & cross-references ─────────────────────────────────────────────
\usepackage[
    colorlinks = true,
    linkcolor  = black,
    citecolor  = black,
    urlcolor   = blue
]{hyperref}

% ── References ────────────────────────────────────────────────────────────────
\usepackage[
    backend  = biber,
    style    = apa,
    sorting  = nyt
]{biblatex}
\addbibresource{references.bib}

% ── Misc ──────────────────────────────────────────────────────────────────────
\usepackage{microtype}
\usepackage{parskip}

% Convenience macro for the information set / conditional-mean notation
\newcommand{\E}{\mathbb{E}}

% ─────────────────────────────────────────────────────────────────────────────
\begin{document}

% ─────────────────────────────────────────────────────────────────────────────
%  Title page
% ─────────────────────────────────────────────────────────────────────────────
\begin{titlepage}
    \centering
    \vspace*{1.2cm}

    {\large IE University} \\[0.25cm]
    {\large Master in Business Analytics \& Data Science} \\[2.6cm]

    {\large\scshape Master's Thesis} \\[1.0cm]

    \rule{\textwidth}{0.4pt} \\[0.65cm]
    {\LARGE\bfseries Mean, Variance, and Market Efficiency \\[0.35cm]
    in the Silver Market\par}
    \vspace{0.65cm}
    {\large\itshape Forecasting Weekly Returns and Realised Volatility with
    Classical, \\[0.15cm] Machine-Learning, and Deep-Learning Models\par}
    \vspace{0.55cm}
    \rule{\textwidth}{0.4pt} \\[2.4cm]

    {\large Asier Ugarteche Perez} \\[1.6cm]

    {\normalsize Supervisor: Prof.\ Dae-Jin Lee} \\[0.4cm]

    \vfill
    {\normalsize Academic Year 2025/2026}
\end{titlepage}
\thispagestyle{empty}

% ── Abstract ──────────────────────────────────────────────────────────────────
\begin{abstract}
\noindent
% DRAFT abstract — qualitative, to be finalised once all results tables are frozen.
This thesis asks how much of silver's short-term price behaviour can be forecast, and
frames the question through the efficient-market hypothesis (EMH). Silver is modelled at
the weekly (Friday-to-Friday) horizon over 2015--2026, and the analysis is factorised
into the two moments: the conditional \emph{mean} of returns and the conditional
\emph{variance}---efficiency constrains the former but not the latter. For the mean, a
deliberate ladder of models of increasing flexibility---ARIMA/ARIMAX, VAR,
mixed-data-sampling (MIDAS) regressions, random forests, gradient boosting, and an
LSTM---is tested against a common random-walk-with-drift benchmark, expanding the
information set from silver's own past (weak-form EMH) to a broad set of public
information (semi-strong-form EMH): cross-asset returns, silver futures positioning,
macroeconomic releases, technical indicators, and sentiment from Reddit and financial
news. For the variance, weekly realised volatility is modelled with GARCH and the
heterogeneous autoregressive (HAR) model, and the question of interest is whether
sentiment improves the forecast. The central finding is a single coherent statement: no
model beats the drift benchmark for the conditional mean by a significant
margin---weak- and semi-strong-form efficiency hold for weekly silver returns---yet the
conditional variance is strongly and significantly forecastable, with sentiment adding a
small but robust increment. Predictable volatility alongside an unpredictable mean is
exactly what the EMH permits, so the two results are complementary rather than
contradictory.
\end{abstract}

\newpage
\thispagestyle{empty}

% ── Table of contents ─────────────────────────────────────────────────────────
\tableofcontents
\newpage

% ═════════════════════════════════════════════════════════════════════════════
\section{Introduction}
\label{sec:intro}
% ═════════════════════════════════════════════════════════════════════════════

Silver is one of the few assets that lives a double life. On one side it is an
industrial input: roughly half of annual demand is consumed in photovoltaics,
electronics, brazing alloys, and medical applications, which ties its price to the
global manufacturing cycle. On the other it is a monetary metal, held as a store of
value, an inflation hedge, and a smaller, more volatile counterpart to gold. Layered on
top of both is a retail-investor following that, in late January 2021, briefly turned
silver into a meme trade: coordinated through Reddit's r/WallStreetSilver---a community
created on 28 January 2021---a wave of retail buying pushed spot from roughly \$25 to
\$30 an ounce within days before fading. This mixture of industrial demand, monetary
hedging, and episodic retail attention makes silver an unusually rich laboratory for
the oldest empirical question in finance: how much of a price can be forecast?

That question has a canonical answer, and it is mostly negative. The efficient-market
hypothesis \parencite{fama1970,fama1991}, building on \textcite{samuelson1965}'s
observation that properly anticipated prices fluctuate randomly, holds that prices
already embed available information, so successive price changes are approximately
unpredictable. The hypothesis is conventionally stated in nested forms graded by the
information set: the \emph{weak} form (past prices and returns carry no exploitable
signal), the \emph{semi-strong} form (no public information does), and the
\emph{strong} form (not even private information does). For a liquid, globally traded
metal, weak- and semi-strong-form efficiency are the natural hypotheses to take to the
data, and they are the two this thesis tests.

A subtlety that is easy to lose is \emph{which} feature of the return distribution
efficiency actually constrains. The hypothesis is a statement about the conditional
\textbf{mean} of returns: the best forecast of next period's return, given the
information available today, is essentially its long-run average. It says nothing about
the conditional \textbf{variance}. Returns can therefore be simultaneously
unpredictable in their mean and highly predictable in their spread---the well-documented
phenomenon of volatility clustering \parencite{engle1982,bollerslev1986}, in which large
moves follow large moves regardless of sign. Efficiency in the first moment and
structure in the second are not in tension. Recognising this is what allows a single
study to say something coherent about both, and it is the organising idea of the thesis.

Empirically, silver's \emph{price} behaves like a random walk: it is non-stationary, and
shocks accumulate rather than mean-revert. The object worth modelling is therefore not
the price level but its increment, the weekly log-return
\(r_t = \log P_t - \log P_{t-1}\). The efficiency question then becomes whether \(r_t\)
can be forecast---first from its own past, and then from the broad set of public
information that a real forecaster would actually possess: cross-asset returns (gold,
the US dollar, copper, equities, oil, and the VIX), futures-market positioning,
macroeconomic releases, technical indicators, and retail and news sentiment.

\textbf{Research questions.} The thesis is built around four questions:
\begin{itemize}
    \item \textbf{RQ1 (weak form).} Are weekly silver returns forecastable from their
    own past?
    \item \textbf{RQ2 (semi-strong form).} Does expanding the information set to public
    covariates help---and does it matter whether the relationship is assumed linear or
    allowed to be nonlinear?
    \item \textbf{RQ3 (the second moment).} Is weekly realised volatility forecastable,
    and if so by how much?
    \item \textbf{RQ4 (sentiment).} Does retail and news sentiment add incremental
    predictive content to either the conditional mean or the conditional variance?
\end{itemize}

\textbf{Approach.} To answer these questions, the analysis works at the weekly
(Friday-to-Friday) horizon and estimates a deliberate ladder of models of increasing
flexibility against a common benchmark and a unified evaluation protocol. The benchmark
throughout is the prevailing-mean ``drift'' forecast---the random-walk-with-drift,
equivalent to ARIMA(0,0,0)---which is the correct null for a return target
\parencite{welch2008,campbell2008}. For the conditional mean, the ladder runs from
classical time-series models (ARIMA/ARIMAX, VAR, and MIDAS regressions) through tree
ensembles (random forests and gradient boosting) to a recurrent neural network (LSTM);
for the conditional variance it runs GARCH and the heterogeneous autoregressive (HAR)
model of \textcite{corsi2009} alongside the same tree ensembles. Forecasts are compared
on out-of-sample \(R^2\) relative to the drift \parencite{campbell2008}, the
Diebold--Mariano test of equal predictive accuracy \parencite{diebold1995}, and---for
the directional reading---the Pesaran--Timmermann test \parencite{pesaran1992}, all on a
held-out 2023--2026 test window.

\textbf{Findings in brief.} The results tell one coherent story. No model---linear or
nonlinear, own-history or public-information---beats the drift benchmark for the
conditional mean by a statistically significant margin; feature-rich and
shorter-window variants do significantly worse. Weak- and semi-strong-form efficiency
hold for weekly silver returns, and they hold more cleanly once the exceptional 2025
bull run is set aside. The conditional variance is a different matter: HAR and GARCH
both forecast weekly realised volatility well clear of the naïve benchmark, and Reddit
attention adds a small but statistically robust increment on top. Crucially, predictable
volatility does \emph{not} contradict efficiency---it is precisely the
mean-versus-variance distinction drawn above---so the two halves of the thesis are not a
success and a failure but two views of the same result: \emph{weekly silver returns are
unforecastable in the mean from past prices and public information, yet their variance
is strongly predictable.} A weak, non-replicating directional signal from futures-market
positioning is the single caveat, and it is best read through the adaptive-markets lens
of \textcite{lo2004} rather than as evidence against efficiency.

\textbf{Contributions.} The thesis makes four contributions. First, it evaluates a full
model ladder---classical, machine-learning, and deep-learning---for \emph{both} moments
of weekly silver returns under one drift benchmark and one significance battery, so that
model families are compared on a level footing rather than across incommensurable
setups. Second, it makes the conditional-mean / conditional-variance factorisation
explicit, turning what could read as ``returns failed, volatility worked'' into a single
coherent efficiency statement. Third, it builds a domain-appropriate sentiment
pipeline---Twitter-RoBERTa for Reddit, FinBERT for financial news---and treats sentiment
as a falsifiable ablation on both moments rather than assuming it helps. Fourth, it
documents a carefully qualified set of negative results on semi-strong predictability of
the mean, separating magnitude from direction and showing where an apparent directional
signal fails to replicate across model classes.

\textbf{Structure.} The remainder of the thesis is organised as follows.
Section~\ref{sec:framework} sets out the conceptual framework---the random-walk view of
prices, the martingale-difference characterisation of the conditional mean, the
expanding information set, and the linear-versus-nonlinear alternatives that motivate the
model ladder. Section~\ref{sec:lit} reviews the relevant literature on commodity-return
predictability, mixed-frequency modelling, realised-volatility forecasting, and financial
sentiment. Section~\ref{sec:data} describes the data and feature construction.
Section~\ref{sec:method} presents the models and the unified evaluation protocol.
Section~\ref{sec:mean} reports the conditional-mean (returns) results, and
Section~\ref{sec:var} the conditional-variance (volatility) results.
Section~\ref{sec:discussion} discusses the findings against the efficiency framing, and
Section~\ref{sec:conclusion} concludes with limitations and directions for future work.

% ═════════════════════════════════════════════════════════════════════════════
\section{Conceptual Framework}
\label{sec:framework}
% ═════════════════════════════════════════════════════════════════════════════

% PLACEMENT NOTE (for the author): this section is the home of the random-walk /
% martingale-difference / information-set notation. The intro states the efficiency
% question in plain English; this section gives it the precise form that motivates the
% model ladder; the Conclusion (sec:conclusion) revisits the one-paragraph synthesis.
% Drafted concisely below; expand per notebooks/weekly/formulas.md as needed.

This section fixes the notation used throughout and states, precisely, the nulls that
each model class is built to test. The aim is to keep the language exact without
overclaiming: the empirical results are framed as tests of the conditional mean and the
conditional variance of weekly silver returns, not as claims about the full return
distribution.

\subsection{Prices, returns, and the random walk}
\label{sec:framework-rw}

Let \(P_t\) be the silver price and \(p_t = \log P_t\) the log price. The weekly
log-return is the increment in the log price,
\begin{equation}
    r_{t+1} = p_{t+1} - p_t, \qquad\text{equivalently}\qquad p_{t+1} = p_t + r_{t+1},
\end{equation}
so the price level and the return are two views of the same process: the level
accumulates shocks, while returns \emph{are} the shocks. A random walk with drift writes
the level as \(p_{t+1} = p_t + \mu + \varepsilon_{t+1}\) with
\(\E[\varepsilon_{t+1}] = 0\). Silver log prices are best described as
\emph{random-walk-like}---non-stationary, with shocks that accumulate---which is the
reason the modelling target throughout is the stationary increment \(r_t\) rather than
the level. The safer wording avoids the \emph{strict} random walk: the innovations are
not independent and identically distributed, because their variance clusters
(Section~\ref{sec:framework-var}).

\subsection{The conditional mean and the information set}
\label{sec:framework-mds}

Let \(F_t\) denote the information available at time \(t\). For the weak-form test this
is silver's own past, \(F_t = \{r_t, r_{t-1}, r_{t-2}, \dots\}\). Returns are a
\emph{martingale difference sequence} (MDS) with respect to \(F_t\) if the conditional
mean equals a constant drift,
\begin{equation}
    \E[r_{t+1} \mid F_t] = \mu,
    \label{eq:weak-null}
\end{equation}
that is, past returns do not move the expected next-week return away from its long-run
average. This is a statement about the first moment \emph{only}; it does not assert that
returns are i.i.d. The semi-strong test expands the information set to public covariates,
\begin{equation}
    G_t = \{F_t,\ \text{gold},\ \text{USD},\ \text{copper},\ \text{VIX},\ \text{oil},\
    \text{COT},\ \text{macro},\ \text{technicals},\ \text{sentiment}\},
\end{equation}
and asks whether the martingale-difference property survives the larger conditioning set,
\begin{equation}
    \E[r_{t+1} \mid G_t] = \mu.
    \label{eq:ss-null}
\end{equation}
ARIMA tests whether silver returns are a martingale difference with respect to their own
past, Equation~\eqref{eq:weak-null}; ARIMAX and the machine-learning models test whether
that property survives once the information set is expanded to public covariates,
Equation~\eqref{eq:ss-null}. Operationally, both nulls are tested as ``does the model
beat the prevailing-mean (drift) benchmark out of sample?''---a forecast that beats the
drift would contradict the corresponding null.

\subsection{Linear and nonlinear alternatives}
\label{sec:framework-nonlin}

The alternative to Equation~\eqref{eq:ss-null} can be specified narrowly or broadly. A
linear model entertains an additive alternative,
\begin{equation}
    \E[r_{t+1} \mid G_t] = \mu + \beta' G_t,
\end{equation}
so ARIMAX, VAR, and MIDAS ask whether public covariates enter the conditional mean
linearly and additively. Nonlinear models entertain the broader alternative
\begin{equation}
    \E[r_{t+1} \mid G_t] = f(G_t),
\end{equation}
where \(f\) may encode thresholds, interactions, regime dependence, and sequence
effects. The motivation is concrete: a signal may be invisible to any single covariate
yet present in a combination. Each condition alone may carry little signal,
\(\E[r_{t+1}\mid \text{high sentiment}] \approx \mu\) and
\(\E[r_{t+1}\mid \text{stretched COT}] \approx \mu\), while the joint condition does not,
\begin{equation}
    \E\!\left[r_{t+1}\mid \text{high sentiment},\ \text{stretched COT},\
    \text{rising VIX},\ \text{falling USD}\right] \neq \mu .
\end{equation}
Random forests, gradient boosting, and the LSTM search this larger space.
The logic of the ladder is therefore cumulative: if even flexible nonlinear models fail
to beat the drift benchmark out of sample, the martingale-difference interpretation is
\emph{strengthened} rather than merely assumed from linear tests. A complementary
own-history check, the generalised spectral test \parencite{hong2003}, probes for
nonlinear departures from Equation~\eqref{eq:weak-null} that linear autocorrelation
diagnostics cannot see.

\subsection{The conditional variance}
\label{sec:framework-var}

Efficiency constrains Equation~\eqref{eq:ss-null}; it places no restriction on the
conditional variance \(\operatorname{Var}(r_{t+1}\mid F_t)\). The two can diverge: the
mean may be (approximately) constant while the variance moves through time, which is
exactly what volatility clustering means---a large move today says little about the sign
or size of next week's return but does signal that next week is likely to be more
volatile. Strict white noise would require independence in \(r_t\), \(r_t^2\), and
\(|r_t|\) alike; the MDS property in Equation~\eqref{eq:weak-null} restricts only the
first of these. Weekly silver returns are close to white noise in the mean but not
strict white noise, because squared returns are strongly autocorrelated. The predictable
structure is therefore concentrated in the second moment, which is what the volatility
chapter (Section~\ref{sec:var}) targets directly via realised volatility.

\subsection{Direction versus magnitude}
\label{sec:framework-direction}

A final distinction governs how forecasts are scored. Directional accuracy asks whether
\(\operatorname{sign}(\hat r_{t+1}) = \operatorname{sign}(r_{t+1})\) more often than
chance, which is \emph{not} the same as beating the drift benchmark under squared-error
loss: a model can call the sign correctly yet fail to improve out-of-sample \(R^2\).
Because efficiency is a conditional-mean claim, the magnitude axis (RMSE, out-of-sample
\(R^2\), and Diebold--Mariano against the drift) is load-bearing, while directional
accuracy---and its significance test, Pesaran--Timmermann \parencite{pesaran1992}---is a
secondary, economically motivated lens. Directional gains are interesting but do not by
themselves reject the martingale-difference null.

% ═════════════════════════════════════════════════════════════════════════════
\section{Literature Review}
\label{sec:lit}
% ═════════════════════════════════════════════════════════════════════════════

% SKELETON — to be written. Planned subsections and the spine of each:

\subsection{Commodity-return predictability and market efficiency}
% Random-walk benchmark hard to beat for commodity spot prices (deaton1992);
% near-unit-root levels vs stationary returns (zhang2010); silver's co-movement with
% gold/USD/equities (batten2010). Anchor the weak/semi-strong framing in fama1970/1991,
% samuelson1965, and the out-of-sample-predictability debate (welch2008, campbell2008).

\subsection{Mixed-frequency models (MIDAS)}
% ghysels2004/2007 framework; Beta vs exponential-Almon weights; andreou2013 on daily
% financial data for lower-frequency targets. Why MIDAS suits macro-at-daily here.

\subsection{Machine learning and deep learning for financial time series}
% Tree ensembles and LSTMs (hochreiter1997); mixed evidence on out-performance
% (sezer2020). Framed as the nonlinear alternative f(G_t) of Section~\ref{sec:framework}.

\subsection{Realised volatility and the HAR model}
% ARCH/GARCH (engle1982, bollerslev1986); HAR-RV (corsi2009); proxy-robust forecast
% evaluation and QLIKE (patton2011). Why predictable variance is consistent with the EMH.

\subsection{Financial sentiment and investor attention}
% FinBERT (araci2019, devlin2019) and Twitter-RoBERTa (loureiro2022, liu2019);
% media-sentiment effects on returns (tetlock2007) and on volatility (audrino2020);
% retail attention and the 2021 squeeze (anand2021, smales2021).

% ═════════════════════════════════════════════════════════════════════════════
\section{Data}
\label{sec:data}
% ═════════════════════════════════════════════════════════════════════════════

% SKELETON — to be written. Planned subsections:
\subsection{Price and cross-asset data}
% Daily silver, gold, USD, copper, S&P 500, VIX, TIPS/real rates, oil; W-FRI weekly
% aggregation; the silver weekly log-return target.
\subsection{Macroeconomic and positioning data}
% FRED monthly macro (CPI, Fed Funds, industrial production, M2) with publication-date
% availability lagging; COT futures positioning.
\subsection{Sentiment data}
% Reddit (r/Silverbugs, r/WallStreetSilver via Arctic Shift) scored with Twitter-RoBERTa;
% GDELT news scored with FinBERT; weekly aggregation and lagging.
\subsection{Feature frame and train/validation/test split}
% Shared weekly W-FRI feature frame; feature groups (EXOG, TECH, MACRO, sentiment, COT);
% chronological split 2015--2021 / 2022 / 2023--2026.

% ═════════════════════════════════════════════════════════════════════════════
\section{Methodology}
\label{sec:method}
% ═════════════════════════════════════════════════════════════════════════════

% SKELETON — to be written. Planned subsections:
\subsection{Forecasting target and benchmarks}
% Weekly log-return target; the drift (prevailing-mean) benchmark = ARIMA(0,0,0);
% the naïve t-1 reference.
\subsection{Conditional-mean models}
% ARIMA/ARIMAX, VAR, MIDAS (monthly-macro and multi-frequency), random forest, XGBoost,
% LSTM (weekly- and daily-input). Walk-forward schemes and the ablation ladder.
\subsection{Conditional-variance models}
% Realised-volatility target; GARCH(1,1); HAR-RV and HAR-X; tree ensembles on HAR+EXOG.
\subsection{Evaluation protocol and significance tests}
% RMSE/MAE, DA/WDA (returns), DCA (volatility); out-of-sample R^2; Diebold--Mariano
% (squared- and absolute-error / QLIKE); Pesaran--Timmermann; ex-2025 robustness.

% ═════════════════════════════════════════════════════════════════════════════
\section{The Conditional Mean: Forecasting Weekly Returns}
\label{sec:mean}
% ═════════════════════════════════════════════════════════════════════════════

% SKELETON — results chapter 1 (returns). Planned subsections:
\subsection{Weak form: own-history models}
% ARIMA(0,0,0) = drift; rolling vs expanding; generalised spectral (MDS) test.
\subsection{Semi-strong form: public information}
% ARIMAX / VAR / MIDAS ladders; RF / XGB / LSTM ladders; OOS R^2 and DM vs drift.
\subsection{Direction versus magnitude}
% DA/WDA, Pesaran--Timmermann; the COT directional caveat and its non-replication.
\subsection{Robustness}
% Absolute-error DM; ex-2025; window schemes.

% ═════════════════════════════════════════════════════════════════════════════
\section{The Conditional Variance: Forecasting Weekly Volatility}
\label{sec:var}
% ═════════════════════════════════════════════════════════════════════════════

% SKELETON — results chapter 2 (volatility). Planned subsections:
\subsection{Realised volatility is forecastable}
% HAR-RV and GARCH vs the naïve RV_{t-1} floor; QLIKE-DM.
\subsection{Cross-asset spillover and nonlinearity}
% HAR+EXOG (linear) and RF/XGB on HAR+EXOG; the cross-asset null.
\subsection{Sentiment and attention}
% HAR+Attention / sentiment ablation; small-but-significant effect; permutation importance.

% ═════════════════════════════════════════════════════════════════════════════
\section{Discussion}
\label{sec:discussion}
% ═════════════════════════════════════════════════════════════════════════════

% SKELETON — to be written. Planned thread:
% Tie the two results chapters back to Section~\ref{sec:framework}: which nulls survived
% (weak + semi-strong on the mean) and which structure is genuine (the variance).
% Reconcile predictable volatility with the EMH (constrains the mean, not the variance).
% The adaptive-markets reading (lo2004) of the COT directional caveat. Economic vs
% statistical significance of the sentiment-volatility effect (audrino2020, tetlock2007).

% ═════════════════════════════════════════════════════════════════════════════
\section{Conclusion}
\label{sec:conclusion}
% ═════════════════════════════════════════════════════════════════════════════

% SKELETON — to be written. The conclusion revisits the one-paragraph synthesis from the
% conceptual framework (Section~\ref{sec:framework} / notebooks/weekly/formulas.md §11):
%
%   Silver log prices are random-walk-like; weekly returns are the appropriate target.
%   Those returns behave as a martingale difference in the conditional mean — own-history
%   models select a constant mean, nonlinear MDS tests fail to reject, and
%   public-information models do not beat the drift out of sample — so weak- and
%   semi-strong-form efficiency hold. Returns are not strict white noise, because squared
%   returns are dependent; the predictable structure lives in the conditional variance,
%   which the volatility chapter forecasts well clear of the naïve floor. One coherent
%   story: the mean is a martingale, the variance is not.
%
% Then: limitations (sample length, sentiment coverage, weekly horizon) and future work
% (intraday data, regime-switching, a VaR capstone tying mean and variance together).

% ─────────────────────────────────────────────────────────────────────────────
%  References
% ─────────────────────────────────────────────────────────────────────────────
\newpage
\printbibliography[heading=bibintoc]

\end{document}
